---
title: "traceR: objective-to-result traceability for CDISC studies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{traceR: objective-to-result traceability for CDISC studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(traceR)
```

## Why traceR

CDISC standards describe a study end to end — the *objective* (USDM), the
*analysis plan* (ARS), the *data* (SDTM, ADaM) and the *results* (ARD) — but the
lineage that connects them lives in separate documents and separate heads.
traceR reads that metadata and assembles a single **traceability graph**: a
typed set of nodes and edges that lets you walk from an objective all the way
down to the SDTM variable that fed a result, and back. Because the graph is
built deterministically from metadata and content-hashed, an AI-assisted
pipeline built on top of it becomes **reproducible by construction**.

## Build a trace graph

The package ships the CDISC pilot metadata as example files. `build_trace()`
reads the USDM study definition, the ARS Reporting Event and the define.xml
files, merges them, and wires the cross-source edges (the endpoint→output
bridge, and ADaM→SDTM `derived_from` lineage).

```{r}
g <- build_trace(
  usdm   = traceR_example("usdm_trace.json"),
  ars    = traceR_example("reporting_event.json"),
  define = list(adam = traceR_example("define_adam.xml"),
                sdtm = traceR_example("define_sdtm.xml"))
)
g
```

A prebuilt copy is also bundled as `cdiscpilot01_trace`.

## Query the lineage

Trace from an objective to a specific output:

```{r}
trace_path(g, "Objective_2", "Out14-3-1-1")[, c("type", "label", "via")]
```

Follow a single relationship — here the predecessor chain of an ADaM variable
back into SDTM:

```{r}
trace_descendants(g, "adam_var:ADAE.TRTAN", rel = "derived_from")[, c("type", "label")]
```

## Validate coverage and find gaps

`trace_coverage()` reports, per objective, how much of the lineage is realised;
`trace_gaps()` lists the structural holes (endpoints with no output, analyses
not yet executed, variables not defined in define.xml, …).

```{r}
trace_coverage(g)[, c("label", "n_outputs", "n_analyses", "reaches_result")]
head(trace_gaps(g), 5)
```

## Execute standard analyses into an ARD

traceR can also *execute* the standard summary analyses straight from the
graph's bindings, emitting the long-skinny ARD keyed to the ARS
`analysis_id`/`operation_id`.

```{r, eval = requireNamespace("cards", quietly = TRUE)}
ex <- execute_reporting_event(g, dirname(traceR_example("adsl.csv")))
head(ex$ard)
```

Merge the ARD back in and the objective now reaches a result:

```{r, eval = requireNamespace("cards", quietly = TRUE)}
g2 <- build_trace(
  usdm = traceR_example("usdm_trace.json"),
  ars  = traceR_example("reporting_event.json"),
  ard  = ex$ard
)
trace_coverage(g2)[, c("label", "n_results", "reaches_result")]
```

## Reproducibility: manifest and verify

`trace_manifest()` writes a content-hashed record of the inputs and the graph;
`trace_verify()` confirms a graph still matches it.

```{r}
man <- trace_manifest(g)
trace_verify(g, man)$reproducible
```

## Export the traceable artifact

`trace_to_json()` emits a portable graph document, and `trace_to_html()`
renders an interactive objective→result explorer (requires visNetwork):

```{r, eval = FALSE}
trace_to_html(g, "traceability.html")
```

## Standards metadata (CDISC Library)

With a CDISC Library member API key, `cdisc_snapshot("sdtmig", "3-4")` downloads
the exhaustive published variable catalogue and caches a versioned snapshot for
reproducibility. Offline, `snapshot_from_define()` derives the same schema from
a study's define.xml. Either snapshot feeds `annotate_standard()`, which flags
each study variable as standard or custom.
```
